---
title: "Changes in Adolescent Alcohol Use"
author: "Group A"
date: "2026-03-07"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(here)
library(tidyverse)
library(tinytex)
library(knitr)
library(kableExtra)
library(nlme)
library(lme4)
library(car)
library(AICcmodavg)
library(patchwork)
library(tibble)
library(gridExtra)
library(grid)

## setting the theme for the plot and table produced, so everything is in
## the same format

theme_set(theme_bw(base_size = 11))

coa_labs <- c(
  "0" = "No COA",
  "1" = "COA"
)

male_labs <- c(
  "0" = "Female",
  "1" = "Male"
)

coa_cols <- c(
  "0" = "#d62728",
  "1" = "#1f77b4"
)


base_plot_theme <- theme(
  plot.title = element_text(size = 11, face = "bold"),
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 10),
  axis.title = element_text(size = 10),
  axis.text = element_text(size = 9)
)

paper_kable <- function(x, caption) {
  
  align_vec <- ifelse(sapply(x, is.numeric), "r", "l")
  
  kable(
    x,
    format = "latex",
    booktabs = TRUE,
    caption = caption,
    align = align_vec,
    linesep = ""
  ) %>%
    kable_styling(
      latex_options = "hold_position",
      font_size = 10
    )
}

if (!dir.exists("images")) dir.create("images")

save_plot_png <- function(plot_obj, filename, width, height, dpi = 300) {
  ggsave(
    filename = file.path("images", filename),
    plot = plot_obj,
    device = "png",
    width = width,
    height = height,
    units = "in",
    dpi = dpi
  )
}

```

## Data Summary

```{r, echo=FALSE}
alc <- read.table("alcohol.txt", header=TRUE) # loading the dataset
alc <- alc %>% 
  mutate(coa = factor(coa, levels=c("0", "1")), # convert COA to a factor
                      male=factor(male, levels=c("0", "1"))) # convert gender to a factor

head(alc)
```

```{r, echo=FALSE}
alc_base <- alc %>% # extract baseline observations (age 14)
  group_by(id) %>%
  slice(1) %>%
  ungroup()

summary_table <- bind_rows(
  
  tibble(
    Characteristic = "Alcohol use",
    Value = ""
  ),
  
  # summarizing alcohol use for each age
  bind_rows(lapply(sort(unique(alc$age)), function(a) { 
    dat <- alc %>% filter(age == a)
    tibble(
      Characteristic = c(
        paste0("\\hspace{1em}Age ", a),
        "\\hspace{2em}Mean (SD)",
        "\\hspace{2em}Median",
        "\\hspace{2em}Range"
      ),
      Value = c(
        "",
        sprintf("%.2f (%.2f)", mean(dat$alcuse, na.rm = TRUE), sd(dat$alcuse, na.rm = TRUE)),
        sprintf("%.2f", median(dat$alcuse, na.rm = TRUE)),
        sprintf("%.2f--%.2f", min(dat$alcuse, na.rm = TRUE), max(dat$alcuse, na.rm = TRUE))
      )
    )
  })),
  
  tibble( # creating a tibble to summarize peer alcohol use
    Characteristic = c(
      "Peer alcohol use",
      "\\hspace{1em}Mean (SD)",
      "\\hspace{1em}Median",
      "\\hspace{1em}Range"
    ),
    Value = c(
      "",
      sprintf("%.2f (%.2f)", mean(alc_base$peer, na.rm = TRUE), sd(alc_base$peer, na.rm = TRUE)),
      sprintf("%.2f", median(alc_base$peer, na.rm = TRUE)),
      sprintf("%.2f--%.2f", min(alc_base$peer, na.rm = TRUE), max(alc_base$peer, na.rm = TRUE))
    )
  ),
  
  # summarize the proportion of each gender
  alc_base %>%
    mutate(
      male_label = case_when(
        male == "0" ~ "Female",
        male == "1" ~ "Male",
        TRUE ~ as.character(male)
      )
    ) %>%
    count(male_label) %>%
    mutate(
      Characteristic = paste0("\\hspace{1em}", male_label),
      Value = paste0(n, " (", round(100 * n / sum(n), 1), "\\%)")
    ) %>%
    select(Characteristic, Value) %>%
    bind_rows(tibble(Characteristic = "Gender", Value = ""), .),
  
  # summarizing the proportion of COA = 0 vs 1
  alc_base %>%
    mutate(
      coa_label = case_when(
        coa == "0" ~ "No",
        coa == "1" ~ "Yes",
        TRUE ~ as.character(coa)
      )
    ) %>%
    count(coa_label) %>%
    mutate(
      Characteristic = paste0("\\hspace{1em}", coa_label),
      Value = paste0(n, " (", round(100 * n / sum(n), 1), "\\%)")
    ) %>%
    select(Characteristic, Value) %>%
    bind_rows(tibble(Characteristic = "COA", Value = ""), .)
)

data_summary_table <- kable(
  summary_table,
  format = "latex",
  booktabs = TRUE,
  caption = "Summary of study variables",
  col.names = c("Variables", "Value"),
  align = "ll",
  linesep = "",
  escape = FALSE
) %>%
  kable_styling(
    latex_options = "hold_position",
    font_size = 10
  )

data_summary_table
```


## Missing data

```{r, echo=FALSE}
missing_table <- alc %>% # summarize missing data by age
  group_by(age) %>%
  summarise(
    "Number of observations" = n(),
    "Alcohol use" = sum(is.na(alcuse)),
    "Peer alcohol use" = sum(is.na(peer)),
    "COA" = sum(is.na(coa)),
    "Gender" = sum(is.na(male))
  )

missing_table
```

## Exploratory plots

```{r fig.width=6, fig.height=4, echo=FALSE}
# creating plots to visualize relations
p_peer_age <- ggplot(alc, aes(x = peer, y = alcuse)) +
  geom_point(alpha = 0.6, size = 2, color = "grey35") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9, color = "steelblue") +
  facet_wrap(~ age) +
  labs(
    x = "Peer alcohol use",
    y = "Alcohol use",
    title = "Figure 1: Association between peer alcohol use and alcohol use by age"
  ) +
  base_plot_theme +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size=10, hjust = 0.5)
  )

p_box_age <- ggplot(alc, aes(x = coa, y = alcuse, fill = coa)) +
  geom_boxplot(alpha = 0.75) +
  facet_wrap(~ age) +
  scale_fill_manual(values = coa_cols, labels = c("No", "Yes")) +
  labs(
    x = "COA",
    y = "Alcohol use",
    fill = "COA",
    title = "Figure 2: Alcohol use by COA across age"
  ) +
  base_plot_theme +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size=10, hjust = 0.5)
  )

p_peer_age_coa <- ggplot(alc, aes(x = peer, y = alcuse, color = coa)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9) +
  facet_wrap(~ age) +
  scale_color_manual(values = coa_cols, labels = c("No", "Yes")) +
  labs(
    x = "Peer alcohol use",
    y = "Alcohol use",
    color = "COA",
    title = "Figure 3: Association between peer alcohol use and alcohol use by age and COA"
  ) +
  base_plot_theme +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size=10, hjust = 0.5)
  )

p_gender <- ggplot(alc, aes(x = age, y = alcuse, color = factor(male))) +
  stat_summary(
    fun = mean,
    geom = "line",
    linewidth = 1.2
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 2
  ) +
  scale_color_manual(
    values = c("grey40", "steelblue"),
    labels = c("Female", "Male"),
    name = "Gender"
  ) +
  labs(
    x = "Age",
    y = "Alcohol use",
    title = "Figure 4. Mean alcohol use by gender across age"
  ) +
  base_plot_theme +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 10, hjust = 0.5)
  )

p_gender_age <- ggplot(alc, aes(x = factor(male), y = alcuse, fill = factor(male))) +
  geom_boxplot(alpha = 0.75) +
  facet_wrap(~ age) +
  scale_fill_manual(
    values = c("grey60", "steelblue"),
    labels = c("Female", "Male"),
    name = "Gender"
  ) +
  labs(
    x = "Gender",
    y = "Alcohol use",
    title = "Figure: Alcohol use by gender across age"
  ) +
  base_plot_theme +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 10, hjust = 0.5)
  )

p_peer_age
p_box_age
p_peer_age_coa
p_gender
p_gender_age
```

### Individual trajectories with mean trend

```{r, fig.width=7, fig.height=5, echo=FALSE}
# plotting the trajectory plot
p_traj <- ggplot(alc, aes(x = age, y = alcuse, group = id, color = coa)) +
  geom_line(alpha = 0.25) +
  stat_summary(
    aes(group = coa, color = coa),
    fun = mean,
    geom = "line",
    linewidth = 1.5
  ) +
  scale_color_manual(
    values = coa_cols,
    labels = c("No", "Yes"),
    name = "COA"
  ) +
  labs(
    x = "Age",
    y = "Alcohol use",
    title = "Figure 5: Alcohol use trajectories by COA"
  ) +
  base_plot_theme +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  )

p_traj
```

## Model comparison

```{r}
# we choose the use linear mixed effect model since we are interested in individual trajectories
# we use the full model to choose the covariance structure
m1 <- lme( # random intercept + ind errors model
  alcuse ~ age * coa * peer + male,
  random = ~ 1 | id,
  data = alc,
  method = "REML"
)

m2 <- lme( # random intercept and slopes + ind errors model
  alcuse ~ age * coa * peer + male,
  random = ~ age | id,
  data = alc,
  method = "REML"
)

m3 <- lme( # random intercept + AR1 errors model
  alcuse ~ age * coa * peer + male,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "REML"
)


m4 <- lme( # random intercept + heteroske ind errors (age) model
  alcuse ~ age * coa * peer + male,
  random = ~ 1 | id,
  weights = varIdent(form = ~ 1 | age),
  data = alc,
  method = "REML"
)

m5 <- lme( # random intercept + heteroske ind errors (gender) model
  alcuse ~ age * coa * peer + male,
  random = ~ 1 | id,
  weights = varIdent(form = ~ 1 | male),
  data = alc,
  method = "REML"
)
```


```{r, echo=FALSE}
cov_table <- tibble(
  `Covariance model` = c(
    "Random intercepts + independent errors",
    "Random intercepts and slopes + independent errors",
    "Random intercepts + AR(1) errors",
    "Random intercepts + heteroskedastic independent errors by age",
    "Random intercepts + heteroskedastic independent errors by gender"
  ),
  AIC = c(AIC(m1), AIC(m2), AIC(m3), AIC(m4), AIC(m5)),
  BIC = c(BIC(m1), BIC(m2), BIC(m3), BIC(m4), BIC(m5))
) %>%
  mutate(
    AIC = round(AIC, 3),
    BIC = round(BIC, 3)
  )

cov_table_print <- paper_kable(cov_table, "Comparison of covariance structures")
cov_table_print
```

Random intercept and slope model has the lowest AIC and BIC.

# Random intercept and slope model

## Fixed effect model selection

```{r}
m_mean1 <- lme( # main effect model
  alcuse ~ age + coa + peer,
  random = ~ age | id,
  data = alc,
  method = "ML"
)

m_mean2 <- lme( # main effect model with gender
  alcuse ~ age + coa + peer + male,
  random = ~ age | id,
  data = alc,
  method = "ML"
)

m_mean3 <- lme( # two way interaction model 
  alcuse ~ age + coa + peer + age:coa + age:peer + coa:peer,
  random = ~ age | id,
  data = alc,
  method = "ML"
)

m_mean4 <- lme( # two way interaction model with gender
  alcuse ~ age + coa + peer + male + age:coa + age:peer + coa:peer,
  random = ~ age | id,
  data = alc,
  method = "ML"
)

m_mean5 <- lme( # three way interaction model
  alcuse ~ age * coa * peer,
  random = ~ age | id,
  data = alc,
  method = "ML"
)

m_mean6 <- lme( # three way interaction model with gender
  alcuse ~ age * coa * peer + male,
  random = ~ age | id,
  data = alc,
  method = "ML"
)
```

```{r, echo=FALSE}
# likelihood ratio tests
lrt_no_gender <- anova(m_mean1, m_mean3, m_mean5) %>%
  as.data.frame()

lrt_with_gender <- anova(m_mean2, m_mean4, m_mean6) %>%
  as.data.frame()

# one combined table
mean_model_compare_all <- bind_rows(
  tibble(
    Model = c(
      "1. Main effects",
      "2. Two-way interactions",
      "3. Three-way interactions"
    ),
    Test = c("", "1 vs 2", "2 vs 3"),
    df = lrt_no_gender$df,
    logLik = lrt_no_gender$logLik,
    `L.Ratio` = lrt_no_gender$`L.Ratio`,
    `p-value` = lrt_no_gender$`p-value`,
    AIC = c(AIC(m_mean1), AIC(m_mean3), AIC(m_mean5)),
    BIC = c(BIC(m_mean1), BIC(m_mean3), BIC(m_mean5))
  ),
  tibble(
    Model = c(
      "4. Main effects (with gender)",
      "5. Two-way interactions (with gender)",
      "6. Three-way interactions (with gender)"
    ),
    Test = c("", "4 vs 5", "5 vs 6"),
    df = lrt_with_gender$df,
    logLik = lrt_with_gender$logLik,
    `L.Ratio` = lrt_with_gender$`L.Ratio`,
    `p-value` = lrt_with_gender$`p-value`,
    AIC = c(AIC(m_mean2), AIC(m_mean4), AIC(m_mean6)),
    BIC = c(BIC(m_mean2), BIC(m_mean4), BIC(m_mean6))
  )
) %>%
  mutate(
    across(c(logLik, `L.Ratio`, `p-value`, AIC, BIC), ~ round(.x, 3))
  )

paper_kable(
  mean_model_compare_all,
  "Fixed-effects structure selection under random intercept and slope model"
)
```


```{r}
final_model_int_slope <- lme( # three way interaction model
  alcuse ~ age * coa * peer,
  random = ~ age | id,
  data = alc,
  method = "ML"
)
```


## Model diagnostic

```{r, fig.width=8, fig.height=6, warning=FALSE, message=FALSE, echo=FALSE}
rhat <- resid(final_model_int_slope, type = "normalized")

condmean <- predict(final_model_int_slope)
y <- alc$alcuse
stg1r <- y - condmean

rtype <- c(rep("Raw", length(stg1r)), rep("Standardized", length(stg1r)))
resid <- c(stg1r, rhat)

stg1resids <- data.frame(
  rtype = rtype,
  resid = as.numeric(resid)
)

p1 <- ggplot(stg1resids, aes(x = resid, fill = rtype)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) +
  labs(
    title = "Histogram of raw and standardized residuals",
    y = "Count",
    x = "Residuals",
    fill = NULL
  ) +
  theme(
    legend.position = c(0.78, 0.78)
  ) +
  base_plot_theme

p2 <- ggplot(data.frame(rhat = rhat), aes(sample = rhat)) +
  stat_qq(shape = 1) +
  stat_qq_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Normal Q-Q plot of standardized residuals",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  base_plot_theme

yhat <- predict(final_model_int_slope)

p3 <- ggplot(data.frame(yhat = yhat, rhat = rhat), aes(y = rhat, x = yhat)) +
  geom_point(shape = 1) +
  labs(
    title = "Standardized residuals versus predicted mean",
    x = "Predicted mean",
    y = "Standardized residual"
  ) +
  base_plot_theme

p4 <- alc %>%
  mutate(rhat = rhat) %>%
  group_by(id) %>%
  mutate(rhat1 = lag(rhat)) %>%
  ggplot(aes(y = rhat, x = rhat1)) +
  geom_point(color = "brown", na.rm = TRUE) +
  labs(
    title = "Lagged residual plot",
    y = "Standardized residual",
    x = "Lagged standardized residual"
  ) +
  base_plot_theme

```

```{r, fig.width=16, fig.height=8, warning=FALSE, message=FALSE, echo=FALSE}
re <- ranef(final_model_int_slope) %>%
  rownames_to_column("id")

b0hat <- re$`(Intercept)`
b1hat <- re$age

p5 <- ggplot(data.frame(b0hat = b0hat), aes(sample = b0hat)) +
  stat_qq(shape = 1) +
  stat_qq_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Normal Q-Q plot of random intercepts",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  base_plot_theme

p6 <- ggplot(data.frame(b1hat = b1hat), aes(sample = b1hat)) +
  stat_qq(shape = 1) +
  stat_qq_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Normal Q-Q plot of random slopes for age",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  base_plot_theme

p7 <- ggplot(data.frame(b0hat = b0hat), aes(x = b0hat)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  labs(
    title = "Histogram of random intercepts",
    x = "Estimates",
    y = "Count"
  ) +
  base_plot_theme

p8 <- ggplot(data.frame(b1hat = b1hat), aes(x = b1hat)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  labs(
    title = "Histogram of random slopes for age",
    x = "Estimates",
    y = "Count"
  ) +
  base_plot_theme

fig7 <- (p1 + p2 + p5 + p6 + p3 + p4 + p7 + p8) +
  plot_layout(ncol = 4) +
  plot_annotation(
    title = "Figure 7: Residual Plots for Random Intercept and Slope Model",
    theme = theme(
      plot.title = element_text(size = 24, face = "bold", hjust = 0.5)
    )
  )

fig7
```

The residual plots show some violations, so we tried the second best model from the random-effects model selection.

# Random intercept only model + AR(1) error

## Fixed effect model selection

```{r}
m_mean1 <- lme( # mean effect model
  alcuse ~ age + coa + peer,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "ML"
)

m_mean2 <- lme( # mean effect model with gender
  alcuse ~ age + coa + peer + male,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "ML"
)

m_mean3 <- lme( # two way interaction model
  alcuse ~ age + coa + peer + age:coa + age:peer + coa:peer,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "ML"
)

m_mean4 <- lme( # two way interaction model with gender
  alcuse ~ age + coa + peer + male + age:coa + age:peer + coa:peer,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "ML"
)

m_mean5 <- lme( # three way interaction model
  alcuse ~ age * coa * peer,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "ML"
)

m_mean6 <- lme( # three way interaction model with gender
  alcuse ~ age * coa * peer + male,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "ML"
)
```


```{r, echo=FALSE}
# likelihood ratio tests
lrt_no_gender <- anova(m_mean1, m_mean3, m_mean5) %>%
  as.data.frame()

lrt_with_gender <- anova(m_mean2, m_mean4, m_mean6) %>%
  as.data.frame()

# one combined table
mean_model_compare_all <- bind_rows(
  tibble(
    Model = c(
      "1. Main effects",
      "2. Two-way interactions",
      "3. Three-way interactions"
    ),
    Test = c("", "1 vs 2", "2 vs 3"),
    df = lrt_no_gender$df,
    logLik = lrt_no_gender$logLik,
    `L.Ratio` = lrt_no_gender$`L.Ratio`,
    `p-value` = lrt_no_gender$`p-value`,
    AIC = c(AIC(m_mean1), AIC(m_mean3), AIC(m_mean5)),
    BIC = c(BIC(m_mean1), BIC(m_mean3), BIC(m_mean5))
  ),
  tibble(
    Model = c(
      "4. Main effects (with gender)",
      "5. Two-way interactions (with gender)",
      "6. Three-way interactions (with gender)"
    ),
    Test = c("", "4 vs 5", "5 vs 6"),
    df = lrt_with_gender$df,
    logLik = lrt_with_gender$logLik,
    `L.Ratio` = lrt_with_gender$`L.Ratio`,
    `p-value` = lrt_with_gender$`p-value`,
    AIC = c(AIC(m_mean2), AIC(m_mean4), AIC(m_mean6)),
    BIC = c(BIC(m_mean2), BIC(m_mean4), BIC(m_mean6))
  )
) %>%
  mutate(
    across(c(logLik, `L.Ratio`, `p-value`, AIC, BIC), ~ round(.x, 3))
  )

paper_kable(
  mean_model_compare_all,
  "Fixed-effects structure selection under AR(1) random intercept model"
)
```

Three way interaction without gender is the best model.

```{r}
final_model_ar1 <- lme( # three way interaction model
  alcuse ~ age * coa * peer,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "ML"
)
```


## Model Diagnostic

```{r, fig.width=16, fig.height=8, warning=FALSE, message=FALSE, echo=FALSE}
rhat <- resid(final_model_ar1, type = "normalized")

condmean <- predict(final_model_ar1)
y <- alc$alcuse
stg1r <- y - condmean

rtype <- c(rep("Raw", length(stg1r)), rep("Standardized", length(stg1r)))
resid <- c(stg1r, rhat)

stg1resids <- data.frame(
  rtype = rtype,
  resid = as.numeric(resid)
)

p1 <- ggplot(stg1resids, aes(x = resid, fill = rtype)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50) +
  labs(
    title = "Histogram of raw and standardized residuals",
    y = "Count",
    x = "Residuals",
    fill = NULL
  ) +
  theme(
    legend.position = c(0.78, 0.78)
  ) +
  base_plot_theme

p2 <- ggplot(data.frame(rhat = rhat), aes(sample = rhat)) +
  stat_qq(shape = 1) +
  stat_qq_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Normal Q-Q plot of standardized residuals",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  base_plot_theme

yhat <- predict(final_model_ar1)

p3 <- ggplot(data.frame(yhat = yhat, rhat = rhat), aes(y = rhat, x = yhat)) +
  geom_point(shape = 1) +
  labs(
    title = "Standardized residuals versus predicted mean",
    x = "Predicted mean",
    y = "Standardized residual"
  ) +
  base_plot_theme

p4 <- alc %>%
  mutate(rhat = rhat) %>%
  group_by(id) %>%
  mutate(rhat1 = lag(rhat)) %>%
  ggplot(aes(y = rhat, x = rhat1)) +
  geom_point(color = "brown", na.rm = TRUE) +
  labs(
    title = "Lagged residual plot",
    y = "Standardized residual",
    x = "Lagged standardized residual"
  ) +
  base_plot_theme

re <- ranef(final_model_ar1) %>%
  rownames_to_column("id")

b0hat <- re$`(Intercept)`

p5 <- ggplot(data.frame(b0hat = b0hat), aes(sample = b0hat)) +
  stat_qq(shape = 1) +
  stat_qq_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "Normal Q-Q plot of random intercepts",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  base_plot_theme

p6 <- ggplot(data.frame(b0hat = b0hat), aes(x = b0hat)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 50, fill = "orange") +
  labs(
    title = "Histogram of random intercepts",
    x = "Estimates",
    y = "Count"
  ) +
  base_plot_theme

fig8 <- (p1 + p2 + p5 +
 p3 + p4 + p6) +
  plot_layout(ncol = 3) +
  plot_annotation(
    title = "Figure 8: Residual Plots for Random Intercept Model",
    theme = theme(
      plot.title = element_text(size = 24, face = "bold", hjust = 0.5)
    )
  )

fig8
```

## Model interpretation

```{r, echo=FALSE}
# print the result of the chosen model: three way interaction model
coef_table <- summary(final_model_ar1)$tTable %>%
  as.data.frame() %>%
  rownames_to_column("Variable") %>%
  rename(
    Estimate = Value,
    SE = Std.Error,
    DF = DF,
    `p-value` = `p-value`
  ) %>%
  mutate(
    `95% CI` = paste0(
      "(",
      round(Estimate - 1.96 * SE, 3),
      ", ",
      round(Estimate + 1.96 * SE, 3),
      ")"
    )
  ) %>%
  mutate(
    across(c(Estimate, SE, DF, `p-value`), ~ round(.x, 3))
  ) %>%
  mutate(
    Variable = dplyr::recode(
      Variable,
      "(Intercept)" = "Constant",
      "age" = "Age",
      "coa1" = "COA",
      "peer" = "Peer alcohol use",
      "age:coa1" = "Age × COA",
      "age:peer" = "Age × peer alcohol use",
      "coa1:peer" = "COA × peer alcohol use",
      "age:coa1:peer" = "Age × COA × peer alcohol use"
    )
  ) %>%
  select(Variable, Estimate, SE, `95% CI`, `p-value`)

coef_table_print <- paper_kable(
  coef_table,
  "Estimated coefficients from the final linear mixed-effects model"
)
coef_table_print
```

## Prediction plot

```{r fig.width=7, fig.height=5, echo=FALSE}
# compute conditional predicted values from the final model
condmean <- predict(final_model_ar1, level = 1)

# compute the mean peer alcohol use
peer0 <- mean(alc$peer, na.rm = TRUE)

# create new data for marginal predictions across age and coa
margdat <- expand.grid(
  age = sort(unique(alc$age)),
  coa = levels(alc$coa),
  peer = peer0
)

# compute marginal predicted mean alcohol use
margdat$margmean <- predict(final_model_ar1, newdata = margdat, level = 0)

# combine predicted values with the original dataset
predalc <- as.data.frame(cbind(alc, condmean))

# plot of predicted alcohol use trajectories
p_pred <- ggplot(
  predalc,
  aes(x = age, y = condmean, group = id)
) +
  geom_line(aes(color = coa), linetype = 3, alpha = 0.5) +
  geom_line(
    data = margdat,
    aes(x = age, y = margmean, color = coa, group = coa),
    linewidth = 1.8,
    inherit.aes = FALSE
  ) +
  scale_color_manual(
    values = coa_cols,
    labels = c("No", "Yes"),
    name = "COA"
  ) +
  labs(
    x = "Age",
    y = "Alcohol use",
    title = "Figure 6: Conditional and marginal alcohol-use trajectories"
  ) +
  base_plot_theme +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 12, hjust = 0.5)
  )

p_pred
```

```{r, echo=FALSE}
save_plot_png(p_peer_age, "EDA_figure1.png", 6, 4)
save_plot_png(p_box_age, "EDA_figure2.png", 6, 4)
save_plot_png(p_peer_age_coa, "EDA_figure3.png", 6, 4)
save_plot_png(p_gender, "EDA_figure4.png", 7, 5)
save_plot_png(p_traj, "EDA_figure5.png", 7, 5)
save_plot_png(p_pred, "prediction_figure6.png", 7, 5)

save_plot_png(fig7, "residual_figure7.png", 16, 8)
save_plot_png(fig8, "residual_figure8.png", 14, 7)
```

```{r}

final_model_ar1 <- lme( # three way interaction model
  alcuse ~ age * coa * peer,
  random = ~ 1 | id,
  correlation = corAR1(form = ~ age | id),
  data = alc,
  method = "ML"
)
# get mean and SD of peer use from your data
peer_mean <- mean(alc$peer, na.rm = TRUE)
peer_sd   <- sd(alc$peer, na.rm = TRUE)

newdat <- expand.grid(
  age = seq(14, 16, by = 0.1),
  coa = c(0, 1),
  peer = c(peer_mean - peer_sd, peer_mean + peer_sd)
)
newdat$pred <- predict(final_model_ar1, newdata = newdat, level=0)
newdat <- newdat %>%
  mutate(
    COA = factor(coa, labels = c("Non-COA", "COA")),
    peer_group = factor(peer,
      labels = c("Low peer use", "High peer use")
    )
  )

ggplot(newdat, aes(x = age, y = pred,
                   color = peer_group,
                   linetype = COA)) +
  geom_line(size = 1.2) +
  labs(x = "Age",
    y = "Predicted alcohol use",
    color = "Peer use",
    linetype = "COA status",
    title="Figure 6: Predicted Alcohol Use Trajectories by COA Status and Peer Use Category", caption="Low peer alcohol use = 0.29, high peer alcohol use = 1.74") +
  base_plot_theme +
  theme(    plot.title = element_text(size = 12, hjust = 0.5))

```
